Binning Strategies and Related Loss for Large Data

### 2013-03-22 Paper.Rnw
fixed paths to charts: figure folder created on the fly, charts in images folder are permanent got rid of multicore calls

\documentclass[letterpaper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{images/}}
\usepackage{epstopdf}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

@@

%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

\begin{document}
\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. The demands of large data require us to re-think strategies of visualizing data. Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
\end{abstract}

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
# setwd("/home/susan/Dropbox/GraphicsGroup/")
setwd("/Users/heike/Dropbox/papers/")

setwd("./dbaccess/NewWriteUp/")

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, echo=F,include=F,eval=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./images/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./images/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./images/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./images/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./images/BinScatter.pdf", width=5, height=5)

@@

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/Overplotting.eps}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingCircles.eps}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingAlpha.eps}}

  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/HexBinning.eps}}
  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/BinScatter.eps}}

	\caption{Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05). Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
  \label{fig:scatter-alpha}
\end{figure}

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F,eval=F,include=F,cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      ggsave("./images/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./images/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
@

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./images/Binning-DataAndBin.eps}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./images/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./images/RandomBinningDataPoints2.eps}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,eval=F>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./images/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An interesting variation which results in statistically nice properties is a random, non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center. 
More formally,
\begin{eqnarray}\label{randbin}
b^r(x) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
  \end{array}\right.
\end{eqnarray}

@@

ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
ggsave("./images/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./Binning-DataVisualNumericalCenter.eps}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./images/FreqLoss.eps}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

#multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
@
\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
\end{figure}

@@

<<LossComparison1, echo=FALSE, eval=T, fig.cap='Spatial Loss increases with bin width for each variable. The random algorithm has higher spatial loss than the standard algorithm.', fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth', fig.align='center', fig.pos='H'>>=

# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)

@@

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE>>=
  print(tab, include.rownames=FALSE)
@
% \end{subsection}

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

@@

%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

\begin{document}
\maketitle

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. The demands of large data require us to re-think strategies of visualizing data. Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
\end{abstract}
%\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}

% <<options-knitr,echo=F,include=F>>=
% require(knitr)
% options(replace.assign=TRUE,width=70)
% opts_chunk$set(fig.path='figure/graphics-', cache.path='cache/cache-', fig.align='center', fig.width=5, fig.height=5, fig.show='hold', cache=TRUE, par=TRUE)
% @

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, echo=F,include=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

@@

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{Overplotting.pdf}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingCircles.pdf}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingAlpha.pdf}}

  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}

	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05). Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
\end{figure}

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F,include=F,cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
@

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints2.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An interesting variation which results in statistically nice properties is a random, non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center. 
More formally,
\begin{eqnarray}\label{randbin}
b^r(x) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
  \end{array}\right.
\end{eqnarray}

@@

ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
@
%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure.pdf}
\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
\end{figure}

@@

\begin{figure}[hbtp]
\centering
<<LossComparison1, echo=FALSE, eval=T, fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth'>>=

@@

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE, dependson='LossTableGraphs'>>=
  print(tab, include.rownames=FALSE)
@
% \end{subsection}
